Formalizing Stack Safety as a Security Property

### 2023-01-27 submission5/intro-v2.tex
Draft rest of intro, checkpoint on new key ideas

The protections offered by stack protection mechanisms are commonly described by giving
concrete examples of attacks that they can prevent---corruption of return
addresses, buffer overflows, use of uninitialized variables, etc.---leaving
a more formal characterization to the reader's intuition.
%
To make these notions more precise, we review a selection of specimens from the
menagerie and begin to consider the difficulties involved in specifying and
verifying the protection they offer.

@@

\paragraph{Capabilities}
%
\rb{(There is plenty of variety here, should we focus on the Cerise-and-friends
  (CHERI-adjacent) work exclusively?)}
%
Capability machines are able to represent and manipulate unforgeable tokens of
authority over specific memory regions using a combination of dedicated
out-of-band memory and a collection of specialized ISA instructions.
%
These hardware capabilities can be used to implement \emph{secure calling
conventions} \ldots

\paragraph{Micro-Policies}
%
\rb{(This partly overlaps with the technical preliminaries)}
%
\ldots

@@

Stack safety is challenging to specify in full generality, because the stack is fundamentally
an architecture-specific construct. (Contrast Azevedo de Amorim et al.'s \cite{DBLP:conf/post/AmorimHP18} work on heap safety,
which is broadly consistent across architectures and language semantics, and therefore amenable
to a very general approach.)
The desired behavior of the stack varies based on ISA, calling convention,
and the features of whichever programming language originated the code.
For instance, the current state-of-the-art stack-safety specification from Georges et al.
\cite{Georges22:TempsDesCerises} supposes a machine with only straightforward call-and-return
control flow, and therefore defines {\it Well-bracketed Control Flow} (WBCF) under the assumption that
a callee should always return to its caller. But, in the presence of tail-call elimination, a
return from a callee to a distant ancestor is perfectly reasonable. Any addition of new features
must be accompanied by tweaks to the model.
%
\rb{This point is extremely important!}

We propose a novel characterization of stack safety using the formal tools of language-based
security~\cite{sabelfeld2003language}, decomposing stack safety into
the {\em integrity} and {\em confidentiality} of the caller’s local state
and the callee's behavior during the callee's execution, plus the control flow protection
of {\em well-bracketed control flow}~\cite{SkorstengaardSTKJFP} (\(\wbcf\)).
\ifexceptions
We also extend WBCF to a setting with exceptions,
in which there may be multiple points at which a callee can ``return.''
\fi
%
\rb{Connect this to the shortcomings of previous work. I'd give at least a quick
  explanation of what the various properties roughly entail in practice (e.g.,
  secrecy of private data, etc.). The comment about the single extension of
  exceptions seems disconnected from the rest.}

This gives us a total of five core properties: \(\wbcf\)
caller integrity (\(\clri\)), caller confidentiality (\(\clrc\)),
callee integrity (\(\clei\)), and callee confidentiality (\(\clec\)).

We are not attempting to create a universal definition of stack safety for all systems.
Rather, we define the stack safety of a single, fairly realistic system, decomposed into
(1) a lightweight model of the system and its security-relevant features, from which we
derive the integrity, confidentiality, and control-flow requirements of each call and
return, (2) the application of that model to a particular target system, and (3)
the formal criteria for integrity, confidentiality, and control flow
to obey the requirements. We consider the criteria in (3) to represent universal definitions
that are parameterized by (1).
%
\rb{Sharpen, contrast the universal definitions (or a framework of parametric
  definitions, say) with the concrete instance we study, again with its interest
  properly motivated.}

We treat stack safety as multiple properties, rather than one multi-faceted property, because
not all enforcement mechanisms attempt to implement all aspects of stack safety. Some focus
exclusively on control flow, some enforce integrity but not confidentiality, and many
focus on the caller over the callee.
%
\rb{Also sharpen after extended discussion?}

@@

\begin{table*}
  \begin{tabular}{| l | l | l | l | l | l | l | l | l |}
    \hline
    Technique       & Approximate & Deferred & High-level & Partial & Well-bracketed & Integrity & Confidentiality & Formal \\
    \hline
    Stack canaries  & Yes         & Yes      & No         & Yes     & Yes            & No        & No              & XXX    \\
    Shadow stacks   & Yes         & Yes      & No         & Yes     & Yes            & No        & No              & XXX    \\
    Bounds checking & No          & No       & Yes        & No      & Yes            & XXX       & XXX             & XXX    \\
    Split stacks    & No          & No       & Yes        & Yes     & Yes            & No        & No              & XXX    \\
    Capabilities    &             &          &            &         &                &           &                 & XXX    \\
    Micro-policies  &             & Yes      &            & Yes     & Yes            & Yes       & Yes             & XXX    \\
    \hline
  \end{tabular}
  \caption{Comparison of Stack Protection Mechanisms}
  \label{tab:relwork}
\end{table*}

@@

@@

>>>>

The protections offered by stack protection mechanisms are commonly described by giving
concrete examples of attacks that they can prevent---corruption of return
addresses, buffer overflows, use of uninitialized variables, etc.---leaving
a more formal characterization to the reader's intuition.
%
To make these notions more precise, we review a selection of specimens from the
menagerie and begin to consider the difficulties involved in specifying and
verifying the protection they offer.
%
For an assessment of these methods in the context of our stack security
framework, see \cref{XXX} below.

@@

\paragraph{Capabilities}
%
\rb{(There is plenty of variety here, should we focus on the Cerise-and-friends
  (CHERI-adjacent) work exclusively?)}
%
Capability machines can represent and manipulate their homonymous
capabilities---unforgeable tokens of authority over specific regions of
memory---using a combination of dedicated out-of-band memory and a collection of
specialized ISA instructions.
%
The most prominent contemporary representative of this long line of work is
CHERI \cite{}, a modern architecture designed to provide efficient fine-grained
memory protection and compartmentalization.
%
Previous work has demonstrated the use of these general principles \rb{in
  simplified models} to implement \emph{secure calling conventions} for the call
stack \cite{}. Those conventions can be seen as combining the advantages and
protections of various families of software-based methods.
%
They are closely related \rb{too strong?} to this work; we discuss them in more
detail in \cref{XXX}.
%
\rb{HardBound is also related to this}

\paragraph{Micro-Policies}
%
\rb{(This partly overlaps with the technical preliminaries, try to offer a quick
  synthesis here.)}
%
Micro-policy machines are a family of tagged architectures that decorate
addressable memory and registers with arbitrary metadata tags, and extend the
processor with a programmable hardware monitor that uses custom rules on
metadata tags to implement security policies of various types.
%
This flexibility can be harnessed to implement a wide array of stack protection policies with relative ease, covering and going beyond the capabilities \rb{word?} of the various software-based techniques \cite{} \rb{as demonstrated by}.
%
They comprise very general forms of \emph{lazy enforcement}, which can tolerate
potential but harmless violations \rb{the system programmer's motto: ``trust me,
  I know what I'm doing''} while still detecting them if and when their effects
become tangible.
%
Micro-policies are one of the centerpieces of our technical development; we
present them in more detail in \cref{XXX}.

@@

As we begin to intuit, the security of the call stack and its contents can be
viewed through many different lenses. Each of these points of view results in a
\emph{unique specification} of the meaning of \emph{stack safety} for a
particular protection mechanism---mostly an informal specification if one is
produced, and almost always left implicit to some degree.
%
Moreover, different protections operate under different threat models, further
complicating their comparison and the assessment of the safeguards they
effectively offer their clients---the programs themselves.
%
On the face of it, it seems perplexing that there is not a \emph{standard
specification} (or set of specifications) for such a fundamental problem.
Instead of reverse engineering an specific enforcement mechanism obtain a
(necessarily approximate) specification, we turn the problem around on its head
and seek more general solutions from first principles.
%
\rb{(Problems/notes: ``backward specs'' not the most useful to their clients,
  not designed with their security needs in mind---goes in the direction of
  program logics---; to reason about security in complex systems an explicit and
  clear spec is preferable; many of the informal specs are at least
  superficially similar and overlapping; hopefully all of them can be understood
  from a small number of fundamental concepts)}

We propose a novel characterization of stack safety using the formal tools of language-based
security~\cite{sabelfeld2003language}, decomposing stack safety into
the {\em integrity} and {\em confidentiality} of the caller’s local state
and the callee's behavior during the callee's execution, plus the control-flow protection
of {\em well-bracketed control flow}~\cite{SkorstengaardSTKJFP} (\(\wbcf\)).
\ifexceptions
We also extend WBCF to a setting with exceptions,
in which there may be multiple points at which a callee can ``return.''
\fi
%
We treat stack safety as multiple properties, rather than one multi-faceted property, because
not all enforcement mechanisms attempt to implement all aspects of stack safety. Some focus
exclusively on control flow, some enforce integrity but not confidentiality, and many
focus on the caller over the callee.
%
\rb{Return to this after extended discussion?}
%
This gives us a total of five core properties: \(\wbcf\),
caller integrity (\(\clri\)), caller confidentiality (\(\clrc\)),
callee integrity (\(\clei\)), and callee confidentiality (\(\clec\)).
%
\rb{Forward references?}

We are not attempting to create a universal definition of stack safety for all systems.
Rather, we define the stack safety of a single, fairly realistic system, decomposed into
(1) a lightweight model of the system and its security-relevant features, from which we
derive the integrity, confidentiality, and control-flow requirements of each call and
return, (2) the application of that model to a particular target system, and (3)
the formal criteria for integrity, confidentiality, and control flow
to obey the requirements. We consider the criteria in (3) to represent universal definitions
that are parameterized by (1).
%
\rb{FIXME: Contrast some universal definitions (or a parametric framework) with
  the concrete instance we study, again with its interest properly motivated.
  (Coming in key ideas?) No ``gold standard''?}

Stack safety is challenging to specify in full generality, because the stack is fundamentally
an architecture-specific construct. (Contrast Azevedo de Amorim et al.'s \cite{DBLP:conf/post/AmorimHP18} work on heap safety,
which is broadly consistent across architectures and language semantics, and therefore amenable
to a very general approach.)
The desired behavior of the stack varies based on ISA, calling convention,
and the features of whichever programming language originated the code.
%
\rb{(Extremely important!)}
%
For instance, the current state-of-the-art stack-safety specification from Georges et al.
\cite{Georges22:TempsDesCerises} supposes a machine with only straightforward call-and-return
control flow, and therefore defines {\it Well-bracketed Control Flow} (WBCF) under the assumption that
a callee should always return to its caller. But, in the presence of tail-call elimination, a
return from a callee to a distant ancestor is perfectly reasonable. Any addition of new features
must be accompanied by tweaks to the model.
%
\rb{(Make this second point later?)}

\rb{*** KEY IDEAS FROM HERE? ***}

@@

\newcommand*{\rot}[1]{\hbox to1em{\hss\rotatebox[origin=br]{-60}{#1}}}
\newcommand*{\tabyes}{\CIRCLE\xspace}
\newcommand*{\tabno}{\Circle\xspace}
\newcommand*{\tabjein}{\RIGHTcircle\xspace}

\begin{table}
\begin{threeparttable}
  \begin{tabular}{l | c c c c | c c c | c}
    \toprule
    \multicolumn{1}{l |}{Mechanism}
    & \multicolumn{4}{l |}{Features}
    & \multicolumn{3}{l |}{Properties}
    & %% Formal criteria
    \\
    \midrule
    %% Technique
    \multicolumn{1}{c}{}
    & \rot{Strict enforcement}
    & \rot{Deferred checking}
    & \rot{Low-level code support}
    & \multicolumn{1}{c}{\rot{Partial program support}}
    & \rot{Well-bracketed control flow}
    & \rot{Data integrity}
    & \multicolumn{1}{c}{\rot{Data confidentiality}}
    & \rot{Formal criteria}
    \\
    \midrule
    Stack canaries  & \tabno   & \tabjein & \tabyes  & \tabyes  & \tabyes  & \tabno   & \tabno   &          \\
    Shadow stacks   & \tabno   & \tabjein &          & \tabyes  & \tabyes  & \tabno   & \tabno   &          \\
    Bounds checking & \tabyes  & \tabno   & \tabno   & \tabno   & \tabyes  &          &          &          \\
    Split stacks    & \tabyes  & \tabno   & \tabno   & \tabyes  & \tabyes  & \tabno   & \tabno   &          \\
    \midrule
    Capabilities    &          & \tabno   &          & \tabyes  &          &          &          & \tabjein \\
    Micro-policies  & \tabyes  & \tabyes  & \tabyes  & \tabyes  & \tabyes  & \tabyes  & \tabyes  & \tabyes  \\
    \bottomrule
  \end{tabular}
  \caption{Comparison of Stack Protection Mechanisms}
  \begin{tablenotes}
  \item \tabyes indicates a protection mechanism possesses the feature
  \item \tabno indicates the mechanism lacks the feature
  \item \tabjein indicates partial or incomplete support
  \item \rb{Complete and check columns, describe. Capitalize?}
  \item \rb{Order? Split int/conf? Deferred/lazy at the end of features}
  \item \rb{2+ lines for capabilities, before and after using our draft?}
  \end{tablenotes}
  \label{tab:relwork}
\end{threeparttable}
\end{table}

@@

CHERI and lightweight formal methods?
Questionably that they are strictly more realistic in the sense that it is very
costly or requires special sophisticated/experimental types of capability not
available/perhaps not very feasible to implement.
Problems of adapting this work to capability machines: security semantics tightly
coupled to regular machine semantics and to a complex calling convention, with
more complex side conditions...

@@

%%%

\rb{Connect this (novel characterization?) to the shortcomings of previous work (later!). I'd give at least a quick
  explanation of what the various properties roughly entail in practice (e.g.,
  secrecy of private data, etc.). The comment about the single extension of
  exceptions (still in intro) seems disconnected from the rest.}
